---
title: "StromDataAnalysis"
author: "Madhavi Pullela"
date: "Saturday, May 23, 2015"
output: html_document
---

###Coursera - Reproducible Research - Peer Assessment 2

U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database analysis

##SYNOPSIS

This report consists in analyzing the NOAA storm database containing data on extreme climate events. This data was collected during the period from 1950 through 2011. The purpose of this analysis is to answer the following two questions:

1.Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


2.Across the United States, which types of events have the greatest economic consequences?


Main conclusions of the study: 1. Tornado is the most hazordous climate event with more than 5600 deaths and 91400 injuries. 2. Floods have caused the most significant economic damage - more than 157 billion USD.

###DATA PROCESSING

####Load the data

We assume the data file is in the working directory

```{r}
rawdata <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header = TRUE, 
    stringsAsFactors = FALSE)
```

####Extract data corresponding to the 48 events as described in the documentation paragraphs 7.1 through 7.48
.Vector of 48 events as defined in the documentation

```{r}
events <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", 
    "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", 
    "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme cold/Wind Chill", 
    "Flash Flood", "Flood", "Freezing", "Frost/Freeze", "Funnel Cloud", "Hail", 
    "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane/Typhoon", 
    "Ice Storm", "Lakeshore Flood", "Lake-Effect Snow", "Lightning", "Marine Hail", 
    "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", 
    "Seiche", "Sleet", "Storm Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", 
    "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", 
    "Wildfire", "Winter Storm", "Winter Weather")
```

Some events are combined events separated with a slash (e.g 'Hurricane/Typhoon'). I will use regular expressions to extract either a combined event (Hurricane/Typhoon) or any part of it (Hurricane or Typhoon)

```{r}
events_regex <- c("Astronomical Low Tide|Low Tide", "Avalanche", "Blizzard", 
    "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", 
    "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme cold/Wind Chill|Extreme Cold|Wind Chill", 
    "Flash Flood", "Flood", "Freezing", "Frost/Freeze|Frost|Freeze", "Funnel Cloud", 
    "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane/Typhoon|Hurricane|Typhoon", 
    "Ice Storm", "Lakeshore Flood", "Lake-Effect Snow", "Lightning", "Marine Hail", 
    "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind|Marine tstm Wind", 
    "Rip Current", "Seiche", "Sleet", "Storm Tide", "Strong Wind", "Thunderstorm Wind|tstm wind", 
    "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", 
    "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
```

The next step is to extract rows corresponding to the event from the documentation I will also choose the columns which are relevant to our analysis:
.EVTYPE Type of event
.FATALITIES Number of fatalities
.INJURIES Number of injuries
.PROPDMG Amount of property damage in orders of magnitude
.PROPDMGEXP Order of magnitude for property damage (e.g. K for thousands)
.CROPDMG Amount of crop damage in orders of magnitude
.PROPDMGEXP Order of magnitude for crop damage (e.g. M for millions)

```{r}
options(scipen = 999)  # force fixed notation of numbers instead of scientific
cleandata <- data.frame(EVTYPE = character(0), FATALITIES = numeric(0), INJURIES = numeric(0), 
    PROPDMG = numeric(0), PROPDMGEXP = character(0), CROPDMG = numeric(0), CROPDMGEXP = character(0))
for (i in 1:length(events)) {
    rows <- rawdata[grep(events_regex[i], ignore.case = TRUE, rawdata$EVTYPE), 
        ]
    rows <- rows[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
        "CROPDMG", "CROPDMGEXP")]
    CLEANNAME <- c(rep(events[i], nrow(rows)))
    rows <- cbind(rows, CLEANNAME)
    cleandata <- rbind(cleandata, rows)
}

```

Take into account the order of magnitude of property and crop damage (H = hundreds, K = thousands, M = millions, B= billions)

#### Converting the letter exponents to integers

```{r}
cleandata[(cleandata$PROPDMGEXP == "K" | cleandata$PROPDMGEXP == "k"), ]$PROPDMGEXP <- 3
cleandata[(cleandata$PROPDMGEXP == "M" | cleandata$PROPDMGEXP == "m"), ]$PROPDMGEXP <- 6
cleandata[(cleandata$PROPDMGEXP == "B" | cleandata$PROPDMGEXP == "b"), ]$PROPDMGEXP <- 9
cleandata[(cleandata$CROPDMGEXP == "K" | cleandata$CROPDMGEXP == "k"), ]$CROPDMGEXP <- 3
cleandata[(cleandata$CROPDMGEXP == "M" | cleandata$CROPDMGEXP == "m"), ]$CROPDMGEXP <- 6
cleandata[(cleandata$CROPDMGEXP == "B" | cleandata$CROPDMGEXP == "b"), ]$CROPDMGEXP <- 9
```

Compute combined economic damage (property damage + crops damage)
#### multiply property and crops damage by 10 raised to the power of the exponent

```{r}
suppressWarnings(cleandata$PROPDMG <- cleandata$PROPDMG * 10^as.numeric(cleandata$PROPDMGEXP))
suppressWarnings(cleandata$CROPDMG <- cleandata$CROPDMG * 10^as.numeric(cleandata$CROPDMGEXP))
```

#### Computing combined economic damage (property damage + crops damage)

```{r}
suppressWarnings(TOTECODMG <- cleandata$PROPDMG + cleandata$CROPDMG)
cleandata <- cbind(cleandata, TOTECODMG)
# delete 'PROPDMGEXP' and 'CROPDMGEXP'columns which have become unnecessary after conversion
cleandata <- cleandata[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG", 
    "CLEANNAME", "TOTECODMG")]
```

At this stage clean data is ready for plotting graphs

##RESULTS

###Question 1 Across the United States, which types of events are most harmful with respect to population health?

Fatalities and injuries
.Aggregate data for fatalities

```{r}

fatalities <- aggregate(FATALITIES ~ CLEANNAME, data = cleandata, FUN = sum)
fatalities <- fatalities[order(fatalities$FATALITIES, decreasing = TRUE), ]
# 5 most harmful causes of fatalities
fatalitiesMax <- fatalities[1:5, ]
print(fatalitiesMax)
```

.Aggregate data for injuries

```{r}
injuries <- aggregate(INJURIES ~ CLEANNAME, data = cleandata, FUN = sum)
injuries <- injuries[order(injuries$INJURIES, decreasing = TRUE), ]
# 5 most harmful causes of injuries
injuriesMax <- injuries[1:5, ]
print(injuriesMax)
```

```{r}
library(ggplot2)
```

Plot the graphs.
I will plot the data for 5 most dangerous events for each type of damage.

# Fatalities
```{r}
ggplot(data = fatalitiesMax, aes(x = fatalitiesMax$CLEANNAME, y = fatalitiesMax$FATALITIES)) + 
    geom_bar(colour = "black", fill = "yellow", stat = "identity") + xlab("Event Type") + 
    ylab("Number of fatalities") + ggtitle("Total number of fatalities, 1950 - 2011") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Injuries
```{r}
ggplot(data = injuriesMax, aes(x = injuriesMax$CLEANNAME, y = fatalitiesMax$FATALITIES)) + 
    geom_bar(colour = "black", fill = "red", stat = "identity") + xlab("Event Type") + 
    ylab("Number of injuries") + ggtitle("Total number of Injuries, 1950 - 2011") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Note: I decided not to compute the total damage consisting of fatalities and injuries (fatalities + injuries) since they have a different order of magnitude (a damage related to 1 death is far greater than a damage related to a light injury, for example). Throughout this report, I always present data relating to fatalities and injuries separately.

### Question 2 Across the United States, which types of events have the greatest economic consequences?

Property and crops combined economic damage
.Aggregate data for property damage

```{r}
propdmg <- aggregate(PROPDMG ~ CLEANNAME, data = cleandata, FUN = sum)
propdmg <- propdmg[order(propdmg$PROPDMG, decreasing = TRUE), ]
# 5 most harmful causes of injuries
propdmgMax <- propdmg[1:5, ]
print(propdmgMax)
```


.Aggregate data for crop damage
```{r}
cropdmg <- aggregate(CROPDMG ~ CLEANNAME, data = cleandata, FUN = sum)
cropdmg <- cropdmg[order(cropdmg$CROPDMG, decreasing = TRUE), ]
# 5 most harmful causes of injuries
cropdmgMax <- cropdmg[1:5, ]
print(cropdmgMax)
```


.Aggregate total economic damage

```{r}
ecodmg <- aggregate(TOTECODMG ~ CLEANNAME, data = cleandata, FUN = sum)
ecodmg <- ecodmg[order(ecodmg$TOTECODMG, decreasing = TRUE), ]
# 5 most harmful causes of property damage
ecodmgMax <- ecodmg[1:5, ]
print(ecodmgMax)
```


Plot the total economic damage graph. Separate information on damage to property and to crops is listed above.
```{r}
# total economic damage (property + crops)
ggplot(data = ecodmgMax, aes(x = ecodmgMax$CLEANNAME, y = ecodmgMax$TOTECODMG/10^9)) + 
    geom_bar(colour = "black", fill = "blue", stat = "identity") + xlab("Event Type") + 
    ylab("Total damage, bln USD") + ggtitle("Total economic damage  1950 - 2011, billions USD") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


Total number of distinct climatic events as per the documentation that were extracted form the data set:

```{r}
nrow(ecodmg)
```

Data for 46 events as per documentation was extracted. Two events were not found.

##CONCLUSIONS

Tornados have caused the greatest number of fatalities - 5,661 and injuries - 91,407 followed by Heat in terms of fatalities 3,138 (9,224 injuries slightly less than Thunderstorm Wind 9,493 injuries which is the second harrmful cause in terms of injuries).

Floods have caused the most significant economic damage 157,764,680,787 USD (combined for property loss and crops damage) followed by Hurricanes and Typhoons - 44,330,000,800 USD
